Expression data from the Agilent G4502A chip (data frozen by the TCGA Pan Cancer Analysis Group) and their corresponding clinical data were consolidated from the Lung Adenocarcinoma cohort from TCGA.

```{r loadData, include=FALSE}
#outputFilePath = file.path(tempdir(), "tcgaAgilentQcedExpressionSet.rbin")
#used = c("syn1571448", "syn1571481")

## FUNCTIONS TO EXTRACT DATA OBJECTS FROM SYNAPSE AND QC
#####
## ANALYST: BRIAN M. BOT
#####

require(synapseClient)
require(rGithubClient)
require(Biobase)
require(limma)
require(corpcor)
require(ggplot2)

## SOURCE IN THE DATA EXTRACTION FUNCTIONS
myRepo <- getRepo(repository="Sage-Bionetworks/CRC-RASness",
                  ref="branch", refName="dev")
sourceRepoFile(myRepo, "functions/utilityFunctions.R")

## CREATE A DIRECTORY FOR PLOTS TO BE UPLOADED TO SYNAPSE
tcgaAgilentDir <- file.path(tempdir(), "tcgaAgilentQC")
dir.create(tcgaAgilentDir)

lungAgilent <- loadTCGAFileFromEntity("syn1571448")

thesePatients <- sapply(strsplit(colnames(lungAgilent), "-", fixed=T), function(x){
  paste(x[1:3], collapse="-")
})
theseFeatures <- rownames(lungAgilent)

if( all(duplicated(thesePatients) == FALSE) ){
  colnames(lungAgilent) <- thesePatients
} else{
  stop("duplicated patients")
}

## CONVERT TO NUMERIC MATRIX
lungAgilent <- apply(lungAgilent, 2, as.numeric)
rownames(lungAgilent) <- theseFeatures
expr <- lungAgilent

## GRAB THE CLINICAL DATA
clin <- loadTCGAFileFromEntity("syn1571481")

## SUBSET TO COHORT OF INTEREST
clin <- clin[colnames(expr), ]

## REMOVE GENES WHERE AT LEAST ONE PATIENT HAS A MISSING VALUE
expr <- expr[ rowSums(is.na(expr)) == 0, ]

## SVD ON EXPRESSION MATRIX -- ASSESS OVERALL STRUCTURE AND POSSIBLE LATENT STRUCTURE
s <- fs(expr)
```

Expression values were already consolidated to the gene level, but were assessed for latent structure in the data by exploring the Singular Value Decomposition (SVD). See below:

```{r svdRaw, echo=FALSE}
qplot(1:length(s$d), s$d,
      xlab="eigen gene",
      ylab="% variance explained")
qplot(s$v[, 1], s$v[, 2],
      xlab="1st svd",
      ylab="2nd svd")
```

Age and gender effects were removed on a gene-by-gene basis. The following plots are SVD plots after age and gender effects were removed to ensure this did not introduce any unwanted effects to the data.

```{r adjustExpression, include=FALSE}
## ADJUST DATA FOR AGE AND GENDER
fit <- lmFit(expr, model.matrix(~as.numeric(clin$age_at_initial_pathologic_diagnosis) + factor(clin$gender)))
exprAdj <- residuals(fit, expr)

## PLOTS FOR ADJUSTED EXPRESSION
sAdj <- fs(exprAdj)
```


```{r svdAdjusted, echo=FALSE}
qplot(1:length(sAdj$d), sAdj$d,
      xlab="eigen gene",
      ylab="% variance explained")
qplot(sAdj$v[, 1], sAdj$v[, 2],
      xlab="1st svd",
      ylab="2nd svd")
```


```{r saveObject, include=FALSE}
tcgaAgilentQcedExpressionSet <- ExpressionSet(exprAdj)
pData(tcgaAgilentQcedExpressionSet) <- clin
save(tcgaAgilentQcedExpressionSet, file=file.path(tempdir(), "tcgaAgilentQcedExpressionSet.rbin"))
```

This resulting R object is an `ExpressionSet` containing expression values on `r nrow(tcgaAgilentQcedExpressionSet)` genes as well as clinical information on `r ncol(tcgaAgilentQcedExpressionSet)` samples within the `phenoData` slot of the object.
