Expression data from the Illumina Hi-seq machines (RNAseqV2 data frozen by the TCGA Pan Cancer Analysis Group) and their corresponding clinical data were consolidated across the Colon Adenocarcinoma (COAD) and Rectum Adenocarcinoma (READ) cohorts from TCGA. Only genes that had at least one sample with at least one count was included in QC processing. Values were transformed by adding one and taking the log base 2 (`log2(expr+1)`).

```{r loadData, include=FALSE}
#outputFilePath = file.path(tempdir(), "tcgaCrcRNAseqQcedExpressionSet.rbin")
#used = c("syn1446197", "syn1446276", "syn1446080", "syn1446153")

## FUNCTIONS TO EXTRACT DATA OBJECTS FROM SYNAPSE AND QC
#####
## ANALYST: BRIAN M. BOT
#####

require(synapseClient)
require(rGithubClient)
require(Biobase)
require(limma)
require(corpcor)
require(ggplot2)

## SOURCE IN THE DATA EXTRACTION FUNCTIONS
myRepo <- getRepo(repository="Sage-Bionetworks/CRC-RASness",
                  ref="branch", refName="dev")
sourceRepoFile(myRepo, "functions/utilityFunctions.R")

## CREATE A DIRECTORY FOR PLOTS TO BE UPLOADED TO SYNAPSE
tcgaRNAseqDir <- file.path(tempdir(), "tcgaRNAseqQC")
dir.create(tcgaRNAseqDir)

coadRNAseq <- loadTCGAFileFromEntity("syn1446197")
readRNAseq <- loadTCGAFileFromEntity("syn1446276")

if( all(rownames(coadRNAseq) == rownames(readRNAseq)) ){
  theseFeatures <- rownames(coadRNAseq)
  crcRNAseq <- cbind(coadRNAseq, readRNAseq)
} else{
  stop("rownames do not match")
}

thesePatients <- sapply(strsplit(colnames(crcRNAseq), "-", fixed=T), function(x){
  paste(x[1:3], collapse="-")
})

if( all(duplicated(thesePatients) == FALSE) ){
  colnames(crcRNAseq) <- thesePatients
} else{
  stop("duplicated patients")
}

## CONVERT TO NUMERIC MATRIX
crcRNAseq <- apply(crcRNAseq, 2, as.numeric)
rownames(crcRNAseq) <- theseFeatures
expr <- crcRNAseq

## GRAB THE CLINICAL DATA
coadClin <- loadTCGAFileFromEntity("syn1446080")
readClin <- loadTCGAFileFromEntity("syn1446153")
coadClin$rns <- rownames(coadClin)
readClin$rns <- rownames(readClin)

crcClin <- merge(x=coadClin, y=readClin, all=T)
rownames(crcClin) <- crcClin$rns
crcClin$rns <- NULL
clin <- crcClin

## SUBSET TO COHORT OF INTEREST
clin <- clin[colnames(expr), ]

## REMOVE GENES WHERE THERE ARE NO COUNTS AVAILABLE FOR ANY PATIENTS
expr <- expr[ rowSums(expr) != 0, ]

## GET RID OF GENES WITH NO GENE SYMBOL
rns <- sapply(strsplit(rownames(expr), "|", fixed=T), "[[", 1)
idx <- rns != "?"
expr <- expr[idx, ]
expr <- log2(expr+1)

## ONE DUPLICATED GENE - COLLAPSE BY USING SVD
expr <- combineProbesToGene(expr, rns[idx])

## SVD ON EXPRESSION MATRIX -- ASSESS OVERALL STRUCTURE AND POSSIBLE LATENT STRUCTURE
s <- fs(expr)
tissue <- factor(ifelse(grepl("RECTUM", clin$tumor_tissue_site), "rectum", "colon"))
```

The few genes that had multiple expression values were consolidated to a single value via the first eigen gene. The subsequent expression values were assessed for latent structure by exploring the Singular Value Decomposition (SVD) of the expression matrix. As can be seen in plots below, there does not appear to be any latent structure.

```{r svdRaw, echo=FALSE}
qplot(1:length(s$d), s$d,
      xlab="eigen gene",
      ylab="% variance explained")
qplot(s$v[, 1], s$v[, 2], colour=tissue,
      xlab="1st svd",
      ylab="2nd svd")
```

Age and gender effects were subsequently removed. The following plots are SVD plots after age and gender effects were removed to ensure this did not introduce any unwanted effects to the data.

```{r adjustExpression, include=FALSE}
## ADJUST DATA FOR AGE AND GENDER
fit <- lmFit(expr, model.matrix(~as.numeric(clin$age_at_initial_pathologic_diagnosis) + factor(clin$gender)))
exprAdj <- residuals(fit, expr)

## PLOTS FOR ADJUSTED EXPRESSION
sAdj <- fs(exprAdj)
```


```{r svdAdjusted, echo=FALSE}
qplot(1:length(sAdj$d), sAdj$d,
      xlab="eigen gene",
      ylab="% variance explained")
qplot(sAdj$v[, 1], sAdj$v[, 2], colour=tissue,
      xlab="1st svd",
      ylab="2nd svd")
```


```{r saveObject, include=FALSE}
tcgaCrcRNAseqQcedExpressionSet <- ExpressionSet(exprAdj)
pData(tcgaCrcRNAseqQcedExpressionSet) <- clin
save(tcgaCrcRNAseqQcedExpressionSet, file=file.path(tempdir(), "tcgaCrcRNAseqQcedExpressionSet.rbin"))
```

This resulting R object is an `ExpressionSet` containing expression values on `r nrow(tcgaCrcRNAseqQcedExpressionSet)` genes as well as clinical information on `r ncol(tcgaCrcRNAseqQcedExpressionSet)` samples within the `phenoData` slot of the object.
