The raw gene expression data for the Julien study was accessed through Array Express through its unique id [E-MTAB-991](http://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-991). These data were generated on XXXXXXX. These data were QCed initially using the `affyPLM` package. The following plot shows the Relative Log Expression (RLE) and Normalized Unscaled Standard Errors (NUSE) plots for all samples.

```{r loadData, include=FALSE}
#outputFilePath = file.path(tempdir(), "julienQcedExpressionSet.rbin")
#used = c("syn1669761", "syn1669763", "syn1669765", "syn1669729")

## SCRIPT TO QC AND PROCESS THE EXPRESSION DATA FROM E-MTAB-991
#####
## ANALYST: BRIAN M. BOT
#####

require(synapseClient)
require(rGithubClient)
require(makecdfenv)
require(affy)
require(affyPLM)
require(limma)
require(corpcor)
require(lattice)
require(hgu133a.db)

options(stringsAsFactors=FALSE)

## SOURCE IN THE DATA EXTRACTION FUNCTIONS
myRepo <- getRepo(repository="Sage-Bionetworks/CRC-RASness",
                  ref="branch", refName="dev")
sourceRepoFile(myRepo, "functions/utilityFunctions.R")

## READ IN RAW DATA FROM SYNAPSE
celFiles1 <- downloadEntity("syn1669761")
celFiles2 <- downloadEntity("syn1669763")
celFiles3 <- downloadEntity("syn1669765")
julienDir <- tempfile(pattern="julienDir")
dir.create(julienDir)
unzip(file.path(celFiles1$cacheDir, celFiles1$files), exdir = julienDir)
unzip(file.path(celFiles2$cacheDir, celFiles2$files), exdir = julienDir)
unzip(file.path(celFiles3$cacheDir, celFiles3$files), exdir = julienDir)
theseFiles <- list.celfiles(julienDir, recursive=T, full.names=T)

julienAffy <- ReadAffy(filenames=theseFiles)

## READ IN CLINICAL DATA
clinFile <- downloadEntity("syn1669729")
julienClin <- read.delim(file.path(clinFile$cacheDir, clinFile$files), as.is=T)
rownames(julienClin) <- julienClin$Array.Data.File
julienClin <- julienClin[ sampleNames(julienAffy), ]
## SCAN DATE NOT AVAILABLE

## PERFORM SOME PRELIMINARY QC
pset <- fitPLM(julienAffy)
```

```{r plmPlots, echo=FALSE}
par(mfrow=c(2,1))
RLE(pset, main="RLE for CRC julien")
NUSE(pset, main="NUSE for CRC julien")
par(mfrow=c(1,1))
```

RLE and NUSE plots do not look overly worrisome.

```{r}
exprSet <- rma(julienAffy)

## COMBINE TO ONE PROBE PER GENE
mapDat <- unlist(as.list(hgu133aSYMBOL))[ featureNames(exprSet) ]
exprGene <- combineProbesToGene(exprs(exprSet), mapDat)
exprs(exprSet) <- exprGene

s <- fs(exprGene)
```

We then ran `rma` which consolidates the feature space down to the probeset level. The first SVD was used to collapse multiple probesets to a single expression value per gene (12,704). Singular Value Decomposition (SVD) was performed on the resulting expression matrix to assess its structure.

```{r svdRaw, echo=FALSE}
xyplot(s$d ~ 1:length(s$d),
       xaxt="n",
       xlab="eigen gene",
       ylab="% variance explained")
xyplot(s$v[,2] ~ s$v[,1], groups=factor(julienClin$Factor.Value.genetic.modification.),
       xlab="1st svd",
       ylab="2nd svd")
```

The first eigen gene quite clearly separates into the mouse / human labels which was then regressed out of the expression matrix. SVD plots below show the structure of the expression matrix after removing those effects.

```{r adjustData, include=FALSE}
fit <- lmFit(exprGene, model.matrix(~ s$v[, 1]))
exprAdj <- residuals(fit, exprGene)
exprs(exprSet) <- exprAdj

## PLOTS FOR ADJUSTED EXPRESSION
sAdj <- fs(exprAdj)
```

```{r svdAdj, echo=FALSE}
xyplot(sAdj$d ~ 1:length(sAdj$d),
       xaxt="n",
       xlab="eigen gene",
       ylab="% variance explained")
xyplot(sAdj$v[,2] ~ sAdj$v[,1],  groups=factor(julienClin$Factor.Value.genetic.modification.),
       xlab="1st svd",
       ylab="2nd svd")
```

After removing the first eigen value, SVD plots seem much more reasonable.

```{r saveObject, include=FALSE}
julienQcedExpressionSet <- exprSet
save("julienQcedExpressionSet", file=file.path(tempdir(), "julienQcedExpressionSet.rbin"))
```

This resulting R object is an `ExpressionSet` containing gene-level expression values as well as clinical information on all samples within the `phenoData` slot of the object.
